---
title: "Final_Model"
author: '520009998'
date: "2024-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# In short, for prediction, it is better to use Log-transformed models to both math and port, stepwise is recommend for interpretability and efficiency.

## Data imports

```{r data imports}
library("dplyr")
library("ggplot2")
library("caret")
library("glmnet")
library("knitr")
# Load the data
d1 <- read.table("student-mat.csv", sep=";", header=TRUE)  # Math dataset
d2 <- read.table("student-por.csv", sep=";", header=TRUE)  # Portuguese dataset

# Clean the data by filtering out rows where G3 is 0
d1 <- d1 %>% filter(G3 != 0)
d2 <- d2 %>% filter(G3 != 0)
```

```{r aic comparson, results='hide'}

full_model_math <- lm(G3 ~ ., data = d1)
full_model_port <- lm(G3 ~ ., data = d2)

# based on the code of ppt.qmd
backward_model_math <- step(full_model_math, direction = "backward")
# based on the code of 'Hannah Report.qmd'
bidirection_model_math <- step(full_model_math, direction = "both")

# based on the code of ppt.qmd
backward_model_port <- step(full_model_port, direction = "backward")
# based on the code of 'Hannah Report.qmd'
bidirection_model_port <- step(full_model_port, direction = "both")

b_m_m <- summary(backward_model_math)
bo_m_m <-summary(bidirection_model_math)
b_m_p <- summary(backward_model_port)
bo_m_p <- summary(bidirection_model_port)
```

```{r table for stepwise and backward comparison}
table_for_stepwise <- data.frame(
  Model = c("Backward_math", "Stepwise_math", "Backward_port", "Stepwise_port"),
  Observations = c(nrow(d1), nrow(d1), nrow(d2), nrow(d2)),
  Rsquare = c(b_m_m$r.squared, bo_m_m$r.squared, b_m_p$r.squared, bo_m_p$r.squared),
  Adjusted_Rsquare = c(b_m_m$adj.r.squared, bo_m_m$adj.r.squared, b_m_p$adj.r.squared, bo_m_p$adj.r.squared),
  AIC = c(AIC(backward_model_math), AIC(bidirection_model_math), AIC(backward_model_port), AIC(bidirection_model_port)),
  Residual_SE = c(sqrt(sum(residuals(backward_model_math)^2) / df.residual(backward_model_math)),
                  sqrt(sum(residuals(bidirection_model_math)^2) / df.residual(bidirection_model_math)),
                  sqrt(sum(residuals(backward_model_port)^2) / df.residual(backward_model_port)),
                  sqrt(sum(residuals(bidirection_model_port)^2) / df.residual(bidirection_model_port))),
  F_statistic = c(
    paste0(round(b_m_m$fstatistic[1], 2), " on ", b_m_m$fstatistic[2], " and ", b_m_m$fstatistic[3], " DF"),
    paste0(round(bo_m_m$fstatistic[1], 2), " on ", bo_m_m$fstatistic[2], " and ", bo_m_m$fstatistic[3], " DF"),
    paste0(round(b_m_p$fstatistic[1], 2), " on ", b_m_p$fstatistic[2], " and ", b_m_p$fstatistic[3], " DF"),
    paste0(round(bo_m_p$fstatistic[1], 2), " on ", bo_m_p$fstatistic[2], " and ", bo_m_p$fstatistic[3], " DF")
  )
)
table_for_stepwise
```
From the AIC, R square value, and Residual SE values, the backward and stepwise models gives the same result to both math and port.
```{r cross_validation_compare_using_10_folds, results='hide'}
# Cross-Validation Setup
train_control <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation

# Perform CV for each model
cv_backward_math <- train(G3 ~ ., data = d1, method = "lm", trControl = train_control)
cv_stepwise_math <- train(G3 ~ ., data = d1, method = "lmStepAIC", trControl = train_control, direction = "both")

cv_backward_port <- train(G3 ~ ., data = d2, method = "lm", trControl = train_control)
cv_stepwise_port <- train(G3 ~ ., data = d2, method = "lmStepAIC", trControl = train_control, direction = "both")

cv_m_b <- summary(cv_backward_math)
cv_m_s <- summary(cv_stepwise_math)
cv_p_b <- summary(cv_backward_port)
cv_p_s <- summary(cv_stepwise_port)
```

```{r table for cross validation}
cv_results <- data.frame(
  Model = c("Backward_math", "Stepwise_math", "Backward_port", "Stepwise_port"),
  RMSE = c(
    mean(cv_backward_math$results$RMSE),
    mean(cv_stepwise_math$results$RMSE),
    mean(cv_backward_port$results$RMSE),
    mean(cv_stepwise_port$results$RMSE)
  ),
  Rsquared = c(
    mean(cv_backward_math$results$Rsquared),
    mean(cv_stepwise_math$results$Rsquared),
    mean(cv_backward_port$results$Rsquared),
    mean(cv_stepwise_port$results$Rsquared)
  ),
  MAE = c(
    mean(cv_backward_math$results$MAE),
    mean(cv_stepwise_math$results$MAE),
    mean(cv_backward_port$results$MAE),
    mean(cv_stepwise_port$results$MAE)
  )
)

cv_results
```

From the cross validation result.

Math:
The stepwise model provides slightly better performance, with a lower RMSE (0.8589 vs. 0.8664) and higher R² (0.9316 vs. 0.9289) compared to the backward model.
MAE is also lower for the stepwise model, indicating better predictive accuracy.

Port:
The backward model offers a slight advantage with lower MAE (0.6768 vs. 0.6759), though the differences are minimal.
The stepwise model's RMSE (0.9259 vs 0.9189) is higher, but both models yield comparable R² values.

```{r linear comparison with cross validation, results='hide'}
normal_model_math <- lm(G3 ~ ., data = d1)
log_model_math <- lm(log(G3) ~ ., data = d1)

normal_model_port <- lm(G3 ~ ., data = d2)
log_model_port <- lm(log(G3) ~ ., data = d2)

cv_normal_math <- train(G3 ~ ., data = d1, method = "lm", trControl = train_control)
cv_log_math <- train(log(G3) ~ ., data = d1, method = "lm", trControl = train_control)

cv_normal_port <- train(G3 ~ ., data = d2, method = "lm", trControl = train_control)
cv_log_port <- train(log(G3) ~ ., data = d2, method = "lm", trControl = train_control)

```

```{r table for linear comparison}
# Extract Model Metrics for Comparison
comparison_results <- data.frame(
  Dataset = c("Math", "Math", "Portuguese", "Portuguese"),
  Model = c("Normal", "Log-Transformed", "Normal", "Log-Transformed"),
  AIC = c(
    AIC(normal_model_math), AIC(log_model_math),
    AIC(normal_model_port), AIC(log_model_port)
  ),
  Rsquared = c(
    summary(normal_model_math)$r.squared,
    summary(log_model_math)$r.squared,
    summary(normal_model_port)$r.squared,
    summary(log_model_port)$r.squared
  ),
  Adjusted_Rsquared = c(
    summary(normal_model_math)$adj.r.squared,
    summary(log_model_math)$adj.r.squared,
    summary(normal_model_port)$adj.r.squared,
    summary(log_model_port)$adj.r.squared
  ),
  RMSE = c(
    mean(cv_normal_math$results$RMSE), mean(cv_log_math$results$RMSE),
    mean(cv_normal_port$results$RMSE), mean(cv_log_port$results$RMSE)
  ),
  MAE = c(
    mean(cv_normal_math$results$MAE), mean(cv_log_math$results$MAE),
    mean(cv_normal_port$results$MAE), mean(cv_log_port$results$MAE)
  )
)

comparison_results
```

## Math Dataset Analysis
Normal Model:

Highest R-squared (0.9435) and adjusted R-squared, indicating it explains the most variance.
However, the AIC is relatively high (909.08), suggesting it is not the most efficient model in terms of parsimony.

Log-Transformed Model:

Lowest AIC (-621.06), indicating the best fit with fewer parameters.
However, the R-squared (0.9087) is lower than the normal model, suggesting it sacrifices variance explanation for predictive accuracy.
Lowest RMSE (0.1025) and MAE (0.0781) make it the most accurate for predictions.
Stepwise Model (Backward):

Slightly lower R-squared than the normal model (0.9389) but achieves a lower AIC (868.79).

## For the Port:
Normal Model:

Highest R-squared (0.8918) and adjusted R-squared, explaining the most variance.
However, it has a high AIC (1729.74), indicating inefficiency.
Log-Transformed Model:

Lowest AIC (-839.13), indicating the best fit.
However, it has a significantly lower R-squared (0.7713), making it less suitable for interpretability.
It has the lowest RMSE (0.1060) and MAE (0.0669), which makes it highly suitable for predictive purposes.
Stepwise Model (Backward):

Slightly lower R-squared than the normal model (0.8882), but achieves a better AIC (1688.46).
Balanced in terms of both variance explanation and parsimony.

# Plot part for each model, in case you need

```{r plot part for log linear}
plot(log_model_math)
plot(log_model_port)
```

```{r plot for normal linear}
plot(normal_model_math)
plot(normal_model_port)
```

```{r plot for stepwise compare}
plot(bidirection_model_math)
plot(bidirection_model_port)
```

```{r plot for backward compare}
plot(backward_model_math)
plot(backward_model_port)
```
